# Term Frequency and TF-IDF {#sec-tfidf}

**Data for this chapter:** `data/vcu_cares_long.csv`

Raw counts favor words that are common everywhere. What we usually want is the opposite: the words that are distinctive of one group of answers compared with the others. **TF-IDF** is the standard way to find them.

```{r}
#| label: setup

source("R/helpers.R")
library(tidytext)

vcu <- load_vcu()

stops <- stop_words |> filter(lexicon == "snowball")
custom_stops <- tibble(word = c("vcu", "student", "students", "like",
                                "get", "just", "really", "also", "can"))

tokens <- vcu |>
  unnest_tokens(word, text) |>
  anti_join(stops, by = "word") |>
  anti_join(custom_stops, by = "word") |>
  filter(!str_detect(word, "^\\d+$"))
```

## The idea

Two quantities, multiplied together.

**Term frequency (tf)** is how common a word is within a document:

$$
\text{tf}(t, d) = \frac{n_{t,d}}{\sum_k n_{k,d}}
$$

the count of word $t$ in document $d$, divided by the total words in $d$.

**Inverse document frequency (idf)** down-weights words that appear in many documents:

$$
\text{idf}(t) = \ln\left(\frac{N}{n_t}\right)
$$

where $N$ is the number of documents and $n_t$ is the number of documents containing $t$. A word in every document has $\ln(1) = 0$, so it contributes nothing. A word in only one document gets the largest idf.

$$
\text{tf-idf}(t, d) = \text{tf}(t, d) \times \text{idf}(t)
$$

High tf-idf means: this word is frequent *here* and rare *elsewhere*. That is exactly "distinctive."

## A tiny example by hand

Three miniature documents make the arithmetic visible.

```{r}
#| label: toy

toy <- tribble(
  ~doc, ~word,
  "A", "housing", "A", "housing", "A", "events",
  "B", "events",  "B", "events",  "B", "parking",
  "C", "events",  "C", "advising"
)

toy |>
  count(doc, word) |>
  bind_tf_idf(word, doc, n)
```

"Events" appears in all three documents, so its idf is $\ln(3/3) = 0$ and its tf-idf is 0 no matter how often it appears. "Housing" appears only in document A, so it gets $\ln(3/1) \approx 1.10$. Work one row out by hand to convince yourself.

## What is a document, here?

TF-IDF compares documents, so the documents you define decide the answer. We will try two.

## Documents = the three questions

If each *question* is one big document, TF-IDF tells us which words are characteristic of strengths versus concerns versus ideas.

```{r}
#| label: fig-tfidf-question
#| fig-cap: "Words most distinctive of each question."
#| fig-height: 6

tfidf_q <- tokens |>
  count(question, word) |>
  bind_tf_idf(word, question, n)

tfidf_q |>
  group_by(question) |>
  slice_max(tf_idf, n = 10, with_ties = FALSE) |>
  ungroup() |>
  ggplot(aes(tf_idf, reorder_within(word, tf_idf, question))) +
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~question, scales = "free") +
  labs(x = "tf-idf", y = NULL)
```

With only three documents, a word that appears in all three gets idf of zero and vanishes. A word that appears in just one question scores high. The results are real but coarse: with $N = 3$, idf can only take a few values. That is a reason to be careful about tiny document counts.

## Documents = student level

A more useful comparison is by who is speaking. Do graduate students talk about different things than undergraduates?

```{r}
#| label: fig-tfidf-level
#| fig-cap: "Words most distinctive of each student level."
#| fig-height: 6

tfidf_level <- tokens |>
  filter(!is.na(level)) |>
  count(level, word) |>
  bind_tf_idf(word, level, n)

tfidf_level |>
  group_by(level) |>
  filter(n >= 5) |>
  slice_max(tf_idf, n = 10, with_ties = FALSE) |>
  ungroup() |>
  ggplot(aes(tf_idf, reorder_within(word, tf_idf, level))) +
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~level, scales = "free") +
  labs(x = "tf-idf", y = NULL)
```

We added `filter(n >= 5)`. Without it, TF-IDF's top words are dominated by words used once, by one student, in one level. That is a known weakness: **rare words get high idf**, even when rarity just means "a typo" or "a name." A minimum count is the standard guard.

::: callout-warning
## Group sizes are very different

There are roughly 1,000 undergraduates and fewer than 20 professional students. TF-IDF is a ratio, so it does not know that the professional group is tiny. Words that rank high for a group of 17 people may reflect two or three students. Treat small-group results as leads to follow up by reading, not as findings.
:::

## Reading the words in context

A word on a chart is stripped of its sentence. Before you believe a distinctive word, look at it in use:

```{r}
#| label: in-context

top_grad <- tfidf_level |>
  filter(level == "Graduate", n >= 5) |>
  slice_max(tf_idf, n = 3, with_ties = FALSE) |>
  pull(word)
top_grad

vcu |>
  filter(level == "Graduate",
         str_detect(str_to_lower(text), top_grad[1])) |>
  slice_head(n = 5) |>
  select(question, text)
```

This is the check we will do after every method in this book. A number tells you where to look; the text tells you what it means.

## TF-IDF as a feature matrix

TF-IDF has a second life. If every answer is a document, then every answer becomes a row of numbers, one per word. That matrix is the input to topic models and to the supervised models in @sec-supervised. Here is the long form `tidytext` produces:

```{r}
#| label: doc-tfidf

doc_tfidf <- tokens |>
  count(doc_id, word) |>
  bind_tf_idf(word, doc_id, n)

doc_tfidf |> arrange(desc(tf_idf)) |> head(10)
```

Notice the top rows. A one-word answer has a term frequency of 1, and a word that appears in a single answer gets the maximum idf, so one-word answers made of rare words dominate. Several of these are typos or keyboard mashing ("asdfasdf"), not findings.

That is a property of the weighting, not a discovery. It is also a handy quality check: sorting by tf-idf is a fast way to find junk answers that your cleaning rules missed.

## Summary

- TF-IDF finds words that distinguish one set of documents from others.
- The documents are your choice, and the answer depends on it.
- Rare words score high; guard against it with a minimum count.
- Small groups give unstable results. Read the underlying text.
