# Topic Models I: LDA {#sec-lda}

**Data for this chapter:** `data/vcu_cares_long.csv`

TF-IDF told us which words are distinctive of a group we had already defined. A **topic model** goes the other way. It proposes the groups. Given only the words in each document, it finds sets of words that tend to appear together and calls each set a topic.

The most widely used topic model is **Latent Dirichlet Allocation (LDA)** (Blei, Ng, & Jordan, 2003).

```{r}
#| label: setup

source("R/helpers.R")
library(tidytext)
library(topicmodels)

vcu <- load_vcu()
```

## The idea in plain terms

LDA imagines that every document was written like this:

1. Each document has its own mix of topics, such as 60% topic A and 40% topic B.
2. Each topic is a probability distribution over words. Topic A might make "housing," "rent," and "dorm" likely.
3. To write each word, pick a topic from the document's mix, then pick a word from that topic.

We observe only the words. LDA works backward to find the topic-word distributions (**beta**) and the document-topic mixes (**gamma**) that would most plausibly produce what we see.

Three things to keep in mind from the start:

- A **topic is a list of words, not a label.** The model does not know that topic 3 is "housing." You read the top words and name it yourself.
- **You choose the number of topics, K.** The data do not tell you the right K.
- **Documents mix topics.** An answer can be partly about advising and partly about money.

## Choosing the documents

LDA learns from words that co-occur *within* a document. Our answers have a median of about a dozen words, and after stop-word removal, six or seven. That is too little for co-occurrence to show up.

The fix is to make the document bigger. We combine each student's answers (strength, concern, and ideas) into one document. Each student's document then has a few dozen words, enough to reveal which words travel together.

```{r}
#| label: docs

stops <- stop_words |> filter(lexicon == "snowball")
custom_stops <- tibble(word = c("vcu", "student", "students", "like",
                                "get", "just", "really", "also", "can",
                                "e", "g", "lot"))

tokens <- vcu |>
  unnest_tokens(word, text) |>
  anti_join(stops, by = "word") |>
  anti_join(custom_stops, by = "word") |>
  filter(!str_detect(word, "^\\d+$"))

# Keep words used in at least 5 different students' answers
vocab <- tokens |>
  distinct(resp_id, word) |>
  count(word) |>
  filter(n >= 5) |>
  pull(word)

student_counts <- tokens |>
  filter(word %in% vocab) |>
  count(resp_id, word)

# Drop students left with too few words to say anything
keep <- student_counts |>
  group_by(resp_id) |>
  filter(sum(n) >= 5) |>
  ungroup()

n_distinct(keep$resp_id); n_distinct(keep$word)
```

Pruning rare words is standard. A word used in only one or two documents cannot help define a topic, and it slows the model down.

## Build the document-term matrix

LDA wants a **document-term matrix**: one row per document, one column per word, cells holding counts.

```{r}
#| label: dtm

dtm <- keep |> cast_dtm(document = resp_id, term = word, value = n)
dtm
```

## Fit the model

```{r}
#| label: fit

K <- 6
lda <- LDA(dtm, k = K, control = list(seed = 664))
```

`seed` matters. LDA starts from a random guess and can land in different places on different runs. Fix the seed so that your results can be reproduced.

## The topics: beta

`tidy(..., matrix = "beta")` returns the probability of each word under each topic. The highest-probability words define the topic.

```{r}
#| label: fig-beta
#| fig-cap: "Top ten words in each topic."
#| fig-height: 7

beta <- tidy(lda, matrix = "beta")

top_terms <- beta |>
  group_by(topic) |>
  slice_max(beta, n = 10, with_ties = FALSE) |>
  ungroup()

ggplot(top_terms, aes(beta, reorder_within(term, beta, topic))) +
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~ topic, scales = "free_y", ncol = 3) +
  labs(x = "Probability of word within topic", y = NULL)
```

**This is the step that requires you.** Read each panel and try to name it. Some will be crisp ("housing, rent, campus, apartments, expensive"). Others will be a grab-bag. A grab-bag topic is not a failure of reading; it is information that K or the preprocessing needs work.

### Words that are common everywhere crowd the lists

A word like "events" or "resources" can be high in several topics simply because it is frequent overall. A useful alternative is to rank words by how much *more* likely they are in one topic than average, which pulls out more distinctive words:

```{r}
#| label: distinctive

distinct_terms <- beta |>
  group_by(term) |>
  mutate(avg_log_beta = mean(log(beta))) |>
  ungroup() |>
  mutate(lift = log(beta) - avg_log_beta) |>
  group_by(topic) |>
  slice_max(lift, n = 8, with_ties = FALSE) |>
  summarise(words = paste(term, collapse = ", "))

distinct_terms
```

Compare these lists with the plot. Distinctive words usually make a topic easier to name; frequent words usually make it easier to defend. Look at both.

## The documents: gamma

`matrix = "gamma"` gives each document's mix of topics.

```{r}
#| label: gamma

gamma <- tidy(lda, matrix = "gamma") |>
  mutate(resp_id = as.integer(document))

gamma |> filter(resp_id == first(resp_id))
```

Each document's gammas sum to 1. How much does a typical student's document belong to a single topic?

```{r}
#| label: fig-dominant
#| fig-cap: "The share of each student's document taken by its strongest topic."

dominant <- gamma |>
  group_by(resp_id) |>
  slice_max(gamma, n = 1, with_ties = FALSE) |>
  ungroup()

ggplot(dominant, aes(gamma)) +
  geom_histogram(binwidth = 0.05, boundary = 0) +
  labs(x = "Gamma of the strongest topic", y = "Students")
```

If the bulk of the mass sits far below 1, documents are genuinely mixed. If it piles up near 1, documents are cleanly single-topic. With short documents you will usually see more mixing than you expect, and the dominant topic can be a weak claim.

## Prevalence: how big is each topic?

```{r}
#| label: fig-prevalence
#| fig-cap: "Average share of each topic across students."

gamma |>
  group_by(topic) |>
  summarise(prevalence = mean(gamma)) |>
  ggplot(aes(prevalence, reorder(factor(topic), prevalence))) +
  geom_col() +
  labs(x = "Mean gamma", y = "Topic")
```

## Always read the documents

The word lists are a summary of a summary. For each topic, read the documents where it is strongest.

```{r}
#| label: read-docs

student_text <- vcu |>
  group_by(resp_id) |>
  summarise(text = paste(text, collapse = " | "), .groups = "drop")

gamma |>
  filter(topic == 1) |>
  slice_max(gamma, n = 3, with_ties = FALSE) |>
  left_join(student_text, by = "resp_id") |>
  select(gamma, text)
```

Change `topic == 1` to each topic in turn. If the top documents for a topic do not look related, the topic is not a theme, no matter how coherent its word list looks.

## A check against the interviewers' codes

We have an independent signal: the topics the interviewers ticked. If LDA is finding real themes, a topic's strongest documents should tend to carry related interviewer codes.

```{r}
#| label: fig-validate
#| fig-cap: "For each LDA topic, the interviewer codes most over-represented among its students."
#| fig-height: 7

codes <- vcu |>
  filter(!is.na(topics)) |>
  separate_rows(topics, sep = ";\\s*") |>
  distinct(resp_id, topics)

base_rate <- codes |>
  count(topics) |>
  mutate(base = n / n_distinct(codes$resp_id)) |>
  select(topics, base)

# Only students from the original form have interviewer codes, so compare
# like with like: the size of each topic is counted among coded students only
topic_sizes <- dominant |>
  filter(resp_id %in% codes$resp_id) |>
  count(topic, name = "n_students")

dominant |>
  inner_join(codes, by = "resp_id") |>
  count(topic, topics) |>
  left_join(topic_sizes, by = "topic") |>
  left_join(base_rate, by = "topics") |>
  mutate(share = n / n_students, lift = share / base) |>
  filter(n >= 5) |>
  group_by(topic) |>
  slice_max(lift, n = 3, with_ties = FALSE) |>
  ungroup() |>
  ggplot(aes(lift, reorder_within(topics, lift, topic))) +
  geom_col() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_y_reordered() +
  facet_wrap(~ topic, scales = "free_y", ncol = 2) +
  labs(x = "Lift over the overall rate (1 = no association)", y = NULL)
```

Lift above 1 means the code is more common among that topic's students than in the data overall. A topic with no codes above 1, or with odd ones, is a candidate for being an artifact.

## How many topics?

There is no right answer, only a trade-off. Fewer topics are broad and easy to read but blur distinct ideas together. More topics are specific but start to split a single theme into near-duplicates.

Fit several values of K and compare them by reading, not just by statistics.

```{r}
#| label: multiple-k

fits <- map(c(4, 6, 8, 10), \(k) LDA(dtm, k = k, control = list(seed = 664)))

map_dfr(fits, \(m) tibble(k = m@k, loglik = as.numeric(logLik(m))))
```

Log-likelihood improves every time you add topics, so it cannot choose K for you. Formal helpers exist: held-out **perplexity** (`topicmodels::perplexity()` on documents the model did not train on) and coherence measures (the `ldatuning` package). They narrow the range, but the last step is always a judgment: does this set of topics tell a story I can defend, and that I can verify by reading?

::: callout-tip
## A practical routine

1. Fit K = 4, 6, 8, 10.
2. For each, print the top words and name every topic.
3. Prefer the smallest K where every topic gets a name you would stand behind and no two topics are near-duplicates.
4. Read top documents for each topic.
:::

## Cautions

- **LDA is stochastic.** Change the seed and the topics reshuffle. Stable themes appear across seeds. Unstable ones are probably noise.
- **Topics are not categories.** A document's strongest topic is a summary, not a classification.
- **Short text is hard.** We combined answers per student to get enough words. Even so, expect messier topics here than with long documents.
- **Preprocessing choices matter.** Different stop-word lists and vocabulary cutoffs will change the topics. Document yours.

In the next chapter, we use a model that lets topic prevalence vary with student characteristics, so that "which topics do graduate students bring up more?" becomes a question the model can answer directly.
