# Topic Models I: LDA {#sec-lda}

**Data for this chapter:** `data/vcu_cares_long.csv`

TF-IDF told us which words are distinctive of a group we had already defined, such as "undergraduates." A **topic model** goes the other way. It proposes the groups. Given only the words in each document, it finds sets of words that tend to appear together and calls each set a **topic**.

The most widely used topic model is **Latent Dirichlet Allocation**, or **LDA** (Blei, Ng, & Jordan, 2003). The name is intimidating, but the idea is not, as the next section shows.

By the end of this chapter you will be able to explain what LDA assumes, fit a model, read its two main outputs (called *beta* and *gamma*), and check whether the topics it found are real.

```{r}
#| label: setup

source("R/helpers.R")
library(tidytext)
library(topicmodels)

vcu <- load_vcu()
```

`topicmodels` is the package that fits the LDA model. `tidytext` (loaded as before) provides helpers that turn the model's output into tables we can work with.

## The idea in plain terms

Imagine you could look inside a pile of answers and see that some are about housing (rent, dorms, repairs), some about money (aid, tuition, costs), and some about advising (advisors, appointments, registration). Nobody labeled them; you just notice that certain words keep turning up together. A topic model does this noticing automatically.

LDA works by imagining a simple story about how every document was written:

1. Each document has its own **mix of topics**, such as 60% topic A and 40% topic B.
2. Each topic is a **list of words with probabilities**. Topic A might make "housing," "rent," and "dorm" likely, and "advisor" unlikely.
3. To write each word of a document, pick a topic according to the document's mix, then pick a word from that topic's list.

Of course, nobody really writes this way. But the story is a useful fiction, because it can be run **backward**. We observe only the words. LDA searches for the topic word-lists and the document mixes that would most plausibly have produced the words we see.

The two things it gives back have Greek-letter nicknames that you will see in all the software:

| Name | What it is | Question it answers |
|---|---|---|
| **beta** (β) | For each topic, how likely each word is | "What is this topic about?" |
| **gamma** (γ) | For each document, how much of it belongs to each topic | "What is this document about?" |

Three things to keep in mind from the start:

- A **topic is a list of words, not a label.** The model does not know that topic 3 is "housing." You read the top words and name it yourself.
- **You choose the number of topics, K.** The data do not tell you the right K.
- **Documents mix topics.** An answer can be partly about advising and partly about money.

## Choosing the documents

LDA learns from words that appear *together within a document*. Our answers have a median of about a dozen words, and after stop-word removal, six or seven. That is too little for patterns of togetherness to show up.

The fix is to make the document bigger. We combine each student's answers (strength, concern, and ideas) into one document. Each student's document then has a few dozen words, enough to reveal which words travel together.

```{r}
#| label: docs

stops <- stop_words |> filter(lexicon == "snowball")
custom_stops <- tibble(word = c("vcu", "student", "students", "like",
                                "get", "just", "really", "also", "can",
                                "e", "g", "lot"))

tokens <- vcu |>
  unnest_tokens(word, text) |>
  anti_join(stops, by = "word") |>
  anti_join(custom_stops, by = "word") |>
  filter(!str_detect(word, "^\\d+$"))

# Keep words used in at least 5 different students' answers
vocab <- tokens |>
  distinct(resp_id, word) |>
  count(word) |>
  filter(n >= 5) |>
  pull(word)

student_counts <- tokens |>
  filter(word %in% vocab) |>
  count(resp_id, word)

# Drop students left with too few words to say anything
keep <- student_counts |>
  group_by(resp_id) |>
  filter(sum(n) >= 5) |>
  ungroup()

n_distinct(keep$resp_id); n_distinct(keep$word)
```

This block does four jobs.

1. **Tokenize and clean** (the first `tokens <-` step). It is the same as in the last two chapters: split into words, remove stop words and bare numbers. (We added "e," "g," and "lot" to the domain stop words because they were noise.)
2. **Build a vocabulary** (`vocab`). `distinct(resp_id, word)` keeps one row per student-word pair, so a word used five times by one student still counts once. `count(word)` then counts how many *different students* used each word, and `filter(n >= 5)` keeps words used by at least five students. `pull(word)` turns the result into a plain list of words. This pruning is standard: a word used by only one or two students cannot help define a topic, and it slows the model down.
3. **Count words per student** (`student_counts`). `filter(word %in% vocab)` keeps only the vocabulary words, and `count(resp_id, word)` counts each word within each student. This is our "documents" table: one row per student-word pair, with a count.
4. **Drop thin documents** (`keep`). `group_by(resp_id)` followed by `filter(sum(n) >= 5)` keeps only students with at least five total words left. Inside a grouped `filter()`, `sum(n)` is calculated within each student, not across the whole table.

The last line reports how many students (documents) and how many distinct words (the vocabulary) remain. `n_distinct()` counts unique values.

## Build the document-term matrix

LDA wants a **document-term matrix**: a grid with one row per document, one column per word, and the count of that word in that document in each cell. It is mostly zeros, because each student uses only a small part of the vocabulary. Such a grid is called *sparse*, and the software stores it compactly.

```{r}
#| label: dtm

dtm <- keep |> cast_dtm(document = resp_id, term = word, value = n)
dtm
```

`cast_dtm()` converts our long table into that grid. Its three options say which column holds the document (`resp_id`), the term (`word`), and the value to put in the cells (`n`, the count). Printing `dtm` shows its size and how sparse it is. "Sparsity" near 98% means 98% of the cells are zero.

## Fit the model

```{r}
#| label: fit

K <- 6
lda <- LDA(dtm, k = K, control = list(seed = 664))
```

- `K <- 6` saves the number of topics under a name, so that we can change it in one place.
- `LDA()` fits the model. The first argument is the matrix; `k` is the number of topics.
- `control = list(seed = 664)` passes extra settings in a list. The `seed` matters: LDA starts from a random guess and can land in different places on different runs. Fixing the seed (see @sec-r-primer) makes your results reproducible.

::: callout-note
## Other options of `LDA()`

- `method`: how the model is fitted. The default, `"VEM"`, is fast. The alternative, `"Gibbs"`, is slower but is often preferred when you want more careful results. Both estimate the same model.
- `control = list(alpha = ...)`: a setting that governs how *mixed* each document is. A small `alpha` pushes each document toward one or two topics; a large one spreads it across many. By default the software estimates it for you, which is a good starting point.
- `control = list(iter = ..., burnin = ...)` (for `"Gibbs"`): how many rounds of computation to run. More rounds give more stable results and take longer.
:::

## The topics: beta

`tidy(lda, matrix = "beta")` turns the fitted model into a table with the probability of each word under each topic. The highest-probability words define the topic.

```{r}
#| label: fig-beta
#| fig-cap: "Top ten words in each topic."
#| fig-height: 7

beta <- tidy(lda, matrix = "beta")

top_terms <- beta |>
  group_by(topic) |>
  slice_max(beta, n = 10, with_ties = FALSE) |>
  ungroup()

ggplot(top_terms, aes(beta, reorder_within(term, beta, topic))) +
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~ topic, scales = "free_y", ncol = 3) +
  labs(x = "Probability of word within topic", y = NULL)
```

- `tidy()` is a general-purpose function that turns the output of a statistical model into a neat table. With `matrix = "beta"` it returns three columns: `topic`, `term` (the word), and `beta` (the probability of that word within that topic).
- The next step keeps the ten most probable words in each topic, the same `group_by()`, `slice_max()`, `ungroup()` recipe as before.
- The chart draws one panel per topic (`facet_wrap(~ topic, ...)`). The option `ncol = 3` arranges the panels in three columns. `scales = "free_y"` lets each panel list its own words.

**This is the step that requires you.** Read each panel and try to name it. Some will be crisp ("housing, rent, campus, apartments, expensive"). Others will be a grab-bag. A grab-bag topic is not a failure of reading; it is information that K or the preprocessing needs work.

### Words that are common everywhere crowd the lists

A word like "events" or "resources" can be high in several topics simply because it is frequent overall. A useful alternative is to rank words by how much *more* likely they are in one topic than average, which pulls out more distinctive words:

```{r}
#| label: distinctive

distinct_terms <- beta |>
  group_by(term) |>
  mutate(avg_log_beta = mean(log(beta))) |>
  ungroup() |>
  mutate(lift = log(beta) - avg_log_beta) |>
  group_by(topic) |>
  slice_max(lift, n = 8, with_ties = FALSE) |>
  summarise(words = paste(term, collapse = ", "))

distinct_terms
```

Reading it in stages:

1. `group_by(term)` then `mutate(avg_log_beta = mean(log(beta)))` computes, for each word, its *average* (log) probability across all the topics. This is the word's "typical" level.
2. `mutate(lift = log(beta) - avg_log_beta)` subtracts that average from the word's value in each topic. The result, `lift`, is large when a word is much more likely in this topic than in a typical one. (Logarithms turn "how many times more likely" into a difference, which is easier to work with.)
3. We then regroup by topic, keep the eight words with the highest lift, and paste them into one line per topic with `paste(term, collapse = ", ")`.

Compare these lists with the plot. Distinctive words usually make a topic easier to name; frequent words usually make it easier to defend. Look at both.

## The documents: gamma

`matrix = "gamma"` gives each document's mix of topics.

```{r}
#| label: gamma

gamma <- tidy(lda, matrix = "gamma") |>
  mutate(resp_id = as.integer(document))

gamma |> filter(resp_id == first(resp_id))
```

The result has one row per document-topic pair: the `document` (the student's ID, stored as text), the `topic`, and the `gamma`. `as.integer(document)` turns the ID back into a number so we can match it with our other tables. The second line shows the six rows for one student: their share of each topic.

Each document's gammas sum to 1. How much does a typical student's document belong to a single topic?

```{r}
#| label: fig-dominant
#| fig-cap: "The share of each student's document taken by its strongest topic."

dominant <- gamma |>
  group_by(resp_id) |>
  slice_max(gamma, n = 1, with_ties = FALSE) |>
  ungroup()

ggplot(dominant, aes(gamma)) +
  geom_histogram(binwidth = 0.05, boundary = 0) +
  labs(x = "Gamma of the strongest topic", y = "Students")
```

For each student (`group_by(resp_id)`), `slice_max(gamma, n = 1)` keeps just their strongest topic. The histogram then shows how big that strongest share usually is.

If the bulk of the bars sits far below 1, documents are genuinely mixed. If it piles up near 1, documents are cleanly single-topic. With short documents you will usually see more mixing than you expect, and "the dominant topic" can be a weak claim.

## Prevalence: how big is each topic?

```{r}
#| label: fig-prevalence
#| fig-cap: "Average share of each topic across students."

gamma |>
  group_by(topic) |>
  summarise(prevalence = mean(gamma)) |>
  ggplot(aes(prevalence, reorder(factor(topic), prevalence))) +
  geom_col() +
  labs(x = "Mean gamma", y = "Topic")
```

`summarise(prevalence = mean(gamma))` averages gamma across all students, separately for each topic. The result is each topic's overall **prevalence**: roughly, the share of all the text that belongs to it. `factor(topic)` tells the chart to treat the topic numbers as labels rather than as quantities.

## Always read the documents

The word lists are a summary of a summary. For each topic, read the documents where it is strongest.

```{r}
#| label: read-docs

student_text <- vcu |>
  group_by(resp_id) |>
  summarise(text = paste(text, collapse = " | "), .groups = "drop")

gamma |>
  filter(topic == 1) |>
  slice_max(gamma, n = 3, with_ties = FALSE) |>
  left_join(student_text, by = "resp_id") |>
  select(gamma, text)
```

- The first part goes back to the original answers and glues each student's answers into one string, separated by `" | "` (`paste(..., collapse = " | ")`). `.groups = "drop"` tells `summarise()` to ungroup afterward.
- The second part picks topic 1, keeps its three strongest students, and uses `left_join()` to attach each student's text. A **left join** adds columns from a second table by matching on a shared column (`resp_id`), keeping every row of the first table.

Change `topic == 1` to each topic in turn. If the top documents for a topic do not look related, the topic is not a theme, no matter how coherent its word list looks.

## A check against the interviewers' codes

We have an independent signal: the topics the interviewers ticked. If LDA is finding real themes, a topic's strongest documents should tend to carry related interviewer codes.

```{r}
#| label: fig-validate
#| fig-cap: "For each LDA topic, the interviewer codes most over-represented among its students."
#| fig-height: 7

codes <- vcu |>
  filter(!is.na(topics)) |>
  separate_rows(topics, sep = ";\\s*") |>
  distinct(resp_id, topics)

base_rate <- codes |>
  count(topics) |>
  mutate(base = n / n_distinct(codes$resp_id)) |>
  select(topics, base)

# Only students from the original form have interviewer codes, so compare
# like with like: the size of each topic is counted among coded students only
topic_sizes <- dominant |>
  filter(resp_id %in% codes$resp_id) |>
  count(topic, name = "n_students")

dominant |>
  inner_join(codes, by = "resp_id") |>
  count(topic, topics) |>
  left_join(topic_sizes, by = "topic") |>
  left_join(base_rate, by = "topics") |>
  mutate(share = n / n_students, lift = share / base) |>
  filter(n >= 5) |>
  group_by(topic) |>
  slice_max(lift, n = 3, with_ties = FALSE) |>
  ungroup() |>
  ggplot(aes(lift, reorder_within(topics, lift, topic))) +
  geom_col() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_y_reordered() +
  facet_wrap(~ topic, scales = "free_y", ncol = 2) +
  labs(x = "Lift over the overall rate (1 = no association)", y = NULL)
```

This is the longest block so far, but it follows one idea: for each LDA topic, which interviewer codes show up *more than usual*?

1. **`codes`**: the interviewers' codes are stored as one text string per answer ("Advising; Safety"). `separate_rows(topics, sep = ";\\s*")` splits that string so each code gets its own row, and `distinct(resp_id, topics)` keeps one row per student-code pair.
2. **`base_rate`**: how common each code is overall, as a share of coded students.
3. **`topic_sizes`**: how many coded students have each LDA topic as their strongest one.
4. **The main pipeline**: `inner_join()` matches each student's dominant topic with their codes (an *inner join* keeps only students who appear in both tables). Then `count(topic, topics)` counts students for each topic-code pair; `left_join()` attaches the topic sizes and base rates; and `mutate()` computes
   - `share`: the fraction of that LDA topic's students who have the code, and
   - `lift`: `share / base`, the ratio of that fraction to the overall rate.
5. `filter(n >= 5)` ignores combinations seen fewer than five times, and `slice_max(lift, n = 3)` keeps the three codes with the highest lift per topic.
6. The chart draws the lifts as bars. `geom_vline(xintercept = 1, linetype = "dashed")` adds a dashed line at 1, where a code is exactly as common as usual.

**Lift** is the key idea. A lift of 2 means the code is twice as common among this topic's students as in the data overall. A lift of 1 means no relationship. A topic with no codes above 1, or with odd ones, is a candidate for being an artifact.

## How many topics?

There is no right answer, only a trade-off. Fewer topics are broad and easy to read but blur distinct ideas together. More topics are specific but start to split a single theme into near-duplicates.

Fit several values of K and compare them by reading, not just by statistics.

```{r}
#| label: multiple-k

fits <- map(c(4, 6, 8, 10), \(k) LDA(dtm, k = k, control = list(seed = 664)))

map_dfr(fits, \(m) tibble(k = m@k, loglik = as.numeric(logLik(m))))
```

This block introduces a way of repeating a step without copying and pasting.

- `map(c(4, 6, 8, 10), \(k) LDA(...))` takes the list of values and runs the function on each one. The odd-looking `\(k) ...` is a tiny function written on the spot ("given `k`, fit an LDA with that many topics"). The result, `fits`, is a list of four fitted models.
- `map_dfr(fits, \(m) tibble(...))` then goes through those models and, for each, builds a one-row table. `m@k` reads the number of topics stored inside the model (the `@` is how R reaches inside this kind of object), and `logLik(m)` is its **log-likelihood**, a score of how well the model fits the data. `map_dfr` stacks the four rows into one table.

Log-likelihood improves every time you add topics, so it cannot choose K for you. Formal helpers exist: held-out **perplexity** (`topicmodels::perplexity()`, which scores the model on documents it did not train on) and coherence measures (the `ldatuning` package). They narrow the range, but the last step is always a judgment: does this set of topics tell a story I can defend, and that I can verify by reading?

::: callout-tip
## A practical routine

1. Fit K = 4, 6, 8, 10.
2. For each, print the top words and name every topic.
3. Prefer the smallest K where every topic gets a name you would stand behind and no two topics are near-duplicates.
4. Read top documents for each topic.
:::

## Cautions

- **LDA is stochastic.** "Stochastic" means it involves randomness. Change the seed and the topics reshuffle. Stable themes appear across seeds. Unstable ones are probably noise.
- **Topics are not categories.** A document's strongest topic is a summary, not a classification.
- **Short text is hard.** We combined answers per student to get enough words. Even so, expect messier topics here than with long documents.
- **Preprocessing choices matter.** Different stop-word lists and vocabulary cutoffs will change the topics. Document yours.

In the next chapter, we use a model that lets topic prevalence vary with student characteristics, so that "which topics do graduate students bring up more?" becomes a question the model can answer directly.

## Summary

- LDA finds **topics** (lists of words that go together) and describes each document as a **mix** of them.
- **Beta** answers "what is this topic about?"; **gamma** answers "what is this document about?"
- You choose **K**, and you name the topics yourself. Both are judgment calls.
- Make documents long enough to show co-occurrence, and prune very rare words.
- Always validate: read the top documents, and compare with any labels you already have.
