# Supervised Learning: Predicting Topic Codes {#sec-supervised}

**Data for this chapter:** `data/vcu_cares_long.csv`

Everything so far has been **unsupervised**: we asked the computer to find patterns without telling it what to look for. In **supervised learning** we give it an answer key. The model learns to reproduce the key from the text, and then we test it on answers it has not seen.

Our answer key is the interviewers' topic codes. For most strength and concern answers in the original form, the interviewer ticked which topics the student discussed (Advising, Dining, Safety, and so on). Can a model look only at the words and predict those ticks?

```{r}
#| label: setup

source("R/helpers.R")
library(tidytext)
library(glmnet)
library(Matrix)

vcu <- load_vcu()
```

## Why bother?

Hand-coding is slow. If a model can reproduce the codes reasonably well, it can code new batches of interviews, flag answers for a second human look, or show which words an interviewer was responding to. The value depends on *how well*, and "how well" has to be measured on data the model never saw.

## Step 1: Build the labeled set

Only answers with at least one topic code can teach or test the model.

```{r}
#| label: labeled

labeled <- vcu |>
  filter(!is.na(topics)) |>
  mutate(
    text_id = row_number(),
    topics = str_split(topics, ";\\s*")
  )

nrow(labeled)
```

Each answer can have several codes, so this is a **multi-label** problem. The easiest way to handle that is to treat each topic as its own yes/no question. Let us see how common each topic is.

```{r}
#| label: fig-topic-freq
#| fig-cap: "How often each topic code was assigned."

topic_counts <- labeled |>
  unnest(topics) |>
  count(topics, sort = TRUE) |>
  mutate(pct = n / nrow(labeled))

topic_counts |>
  ggplot(aes(n, reorder(topics, n))) +
  geom_col() +
  labs(x = "Answers with this code", y = NULL)
```

Some topics are common ("Student activities"), and some are rare ("Staff development"). Rare topics will be hard to learn: with only a dozen examples, there is little for the model to generalize from. We start with one topic and then scale up.

```{r}
#| label: target

target <- "Student activities"
labeled <- labeled |>
  mutate(y = map_lgl(topics, \(t) target %in% t))

mean(labeled$y)
```

About one answer in five carries this code. That imbalance matters for how we judge the model, as we will see.

## Step 2: Split before you do anything else

We will hold back 20% of the data to test the model. Two rules:

1. **Split first.** Anything the model "learns" from the test set, even the vocabulary, makes the test score optimistic. This is called **data leakage**.
2. **Split by student, not by answer.** One student can have a strength answer and a concern answer. If one lands in training and the other in testing, the model has effectively seen the student. Splitting by `resp_id` keeps each student on one side.

```{r}
#| label: split

set.seed(664)
students <- unique(labeled$resp_id)
test_students <- sample(students, size = round(0.2 * length(students)))

train <- labeled |> filter(!resp_id %in% test_students)
test  <- labeled |> filter( resp_id %in% test_students)

c(train = nrow(train), test = nrow(test))
```

## Step 3: Turn text into numbers

The model needs a matrix: one row per answer, one column per word. We use TF-IDF weights from @sec-tfidf. The key discipline is that the **vocabulary and the idf values come only from the training data**, and the test set is converted using those same values.

```{r}
#| label: features

stops <- stop_words |> filter(lexicon == "snowball")

tokenize <- function(d) {
  d |>
    select(text_id, text) |>
    unnest_tokens(word, text) |>
    anti_join(stops, by = "word") |>
    filter(!str_detect(word, "^\\d+$"))
}

train_tok <- tokenize(train)

# Vocabulary: words that appear in at least 5 training answers
vocab_df <- train_tok |>
  distinct(text_id, word) |>
  count(word, name = "df") |>
  filter(df >= 5) |>
  mutate(idf = log(nrow(train) / df))

length(vocab_df$word)

# Convert any set of answers to a TF-IDF matrix over the training vocabulary
featurize <- function(d) {
  tok <- tokenize(d) |>
    count(text_id, word) |>
    inner_join(vocab_df, by = "word") |>
    group_by(text_id) |>
    mutate(tf = n / sum(n)) |>
    ungroup()

  sparseMatrix(
    i = match(tok$text_id, d$text_id),
    j = match(tok$word, vocab_df$word),
    x = tok$tf * tok$idf,
    dims = c(nrow(d), nrow(vocab_df)),
    dimnames = list(NULL, vocab_df$word)
  )
}

x_train <- featurize(train)
x_test  <- featurize(test)
dim(x_train); dim(x_test)
```

Both matrices have the same columns, which is what lets a model trained on one score the other. Words in the test set that never appeared in training are simply ignored: the model has nothing to say about them.

The matrix is **sparse**: nearly every cell is zero, since each answer uses only a handful of the few hundred vocabulary words. Storing it as a sparse matrix keeps it small.

## Step 4: Fit a regularized logistic regression

With more columns than a regression can handle comfortably, we use the **lasso** (`glmnet` with `alpha = 1`). It is logistic regression with a penalty on the size of the coefficients, which pushes most of them exactly to zero. The result is a short list of words that matter. `cv.glmnet()` picks the penalty strength (`lambda`) by cross-validation inside the training data.

```{r}
#| label: fit

set.seed(664)
fit <- cv.glmnet(x_train, train$y, family = "binomial",
                 alpha = 1, type.measure = "auc")

fit$lambda.1se
sum(coef(fit, s = "lambda.1se")[-1] != 0)   # words with a nonzero coefficient
```

`lambda.1se` is the most-penalized model within one standard error of the best. It is simpler than the single best model and usually generalizes about as well.

## Step 5: Judge it on data it has not seen

```{r}
#| label: predict

test$p <- as.numeric(predict(fit, x_test, s = "lambda.1se", type = "response"))

head(test |> select(text, y, p), 6)
```

Each answer gets a predicted probability. To turn it into a yes/no call, pick a threshold. Start at 0.5, and note that this is a choice.

```{r}
#| label: metrics

metrics <- function(truth, p, threshold = 0.5) {
  pred <- p >= threshold
  tp <- sum(pred & truth); fp <- sum(pred & !truth)
  fn <- sum(!pred & truth); tn <- sum(!pred & !truth)
  precision <- if (tp + fp > 0) tp / (tp + fp) else NA_real_
  recall    <- tp / (tp + fn)
  tibble(tp, fp, fn, tn,
         accuracy  = (tp + tn) / length(truth),
         precision, recall,
         f1 = 2 * precision * recall / (precision + recall))
}

# Area under the ROC curve: the chance a random positive outranks a random negative
auc <- function(truth, p) {
  r <- rank(p)
  n1 <- sum(truth); n0 <- sum(!truth)
  (sum(r[truth]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

metrics(test$y, test$p)
auc(test$y, test$p)
```

Read the output before the definitions. At a threshold of 0.5 the model is cautious: when it says "Student activities" it is nearly always right (high precision), but it says so rarely and misses most of the real ones (low recall). The AUC of about 0.8 says the probabilities do carry useful signal; the 0.5 cutoff is just not the best place to draw the line.

Four numbers deserve plain definitions:

- **Accuracy**: the share of all calls that were right.
- **Precision**: of the answers the model called "Student activities," the share that really are.
- **Recall**: of the answers that really are "Student activities," the share the model found.
- **F1**: a single number balancing precision and recall.

### Accuracy alone is misleading

A model that says "no" to every answer gets an accuracy equal to the share of "no" answers:

```{r}
#| label: baseline

1 - mean(test$y)
```

That is the bar to beat. Our model's accuracy is only a few points above it, even though the model is clearly doing something useful, which is why accuracy is a poor yardstick when one class is rare. Precision, recall, and AUC tell you much more. AUC does not depend on a threshold: 0.5 means guessing, 1.0 means perfect.

### The threshold is a dial

Lowering the threshold finds more true answers (recall rises) but also flags more wrong ones (precision falls).

```{r}
#| label: fig-threshold
#| fig-cap: "Precision, recall, and F1 as the threshold changes."

map_dfr(seq(0.05, 0.8, by = 0.05),
        \(th) metrics(test$y, test$p, th) |> mutate(threshold = th)) |>
  select(threshold, precision, recall, f1) |>
  pivot_longer(-threshold) |>
  ggplot(aes(threshold, value, color = name)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c(f1 = vcu_dark, precision = vcu_teal, recall = vcu_gold)) +
  labs(x = "Threshold", y = NULL, color = NULL)
```

Which threshold is right depends on the job. If a human will review every flagged answer, favor recall: you do not want to miss any. If the model's calls go straight into a report, favor precision.

## Step 6: What did it learn?

The lasso gives us a readable answer. The words with the largest positive coefficients push toward "Student activities"; the most negative push away.

```{r}
#| label: fig-coefs
#| fig-cap: "Words most associated with the Student activities code."
#| fig-height: 5.5

coefs <- coef(fit, s = "lambda.1se")
coef_df <- tibble(word = rownames(coefs), beta = as.numeric(coefs)) |>
  filter(word != "(Intercept)", beta != 0)

coef_df |>
  slice_max(abs(beta), n = 24) |>
  ggplot(aes(beta, reorder(word, beta), fill = beta > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c(`TRUE` = vcu_teal, `FALSE` = vcu_gold)) +
  labs(x = "Coefficient (log-odds)", y = NULL)
```

Do these words make sense for the topic? If you see "events," "clubs," and "organizations" pushing up, the model learned something real. If you see odd words, such as the name of one interviewer's favorite example, it is learning quirks of the data, and you should be skeptical.

## Step 7: Read the mistakes

The most useful thing you can do with a classifier is read where it is wrong.

```{r}
#| label: errors

# Missed: coded as the topic, but given the lowest probabilities
test |>
  filter(y) |>
  arrange(p) |>
  slice_head(n = 6) |>
  select(p, text)

# False alarms: not coded as the topic, but given the highest probabilities
test |>
  filter(!y) |>
  arrange(desc(p)) |>
  slice_head(n = 6) |>
  select(p, text)
```

Mistakes usually fall into three bins. Some are real model failures: the answer is clearly about activities, and the model missed it, often because it uses a word the training data never saw. Some are **label noise**: the answer is not really about activities, and the interviewer ticked the box anyway, or the reverse. And some are judgment calls that two human coders would disagree about. If many errors are the second and third kind, the model's ceiling is limited by the labels, not by the algorithm.

The lists above show examples of each. Among the missed answers, several barely mention the topic they were coded with: the code is there, but the words are not. That is label noise, and the model cannot be blamed for failing to find words that are not in the text. Among the false alarms, many are about networking and professional events, which are *near* the Student activities topic but were coded under professional development instead. Whether they belong is a coding-scheme question that no algorithm can settle.

## Step 8: Scaling up to many topics

The same recipe works for any topic. We wrap it in a function and run it for the topics with enough examples. Because we reuse the vocabulary and the split, the comparison across topics is fair.

```{r}
#| label: all-topics
#| cache: false

fit_topic <- function(topic) {
  y_tr <- map_lgl(train$topics, \(t) topic %in% t)
  y_te <- map_lgl(test$topics,  \(t) topic %in% t)

  set.seed(664)
  f <- cv.glmnet(x_train, y_tr, family = "binomial", alpha = 1, type.measure = "auc")
  p <- as.numeric(predict(f, x_test, s = "lambda.1se", type = "response"))

  # Use each topic's own base rate as a data-driven threshold
  th <- mean(y_tr)
  bind_cols(
    tibble(topic = topic, prevalence = mean(y_te), auc = auc(y_te, p)),
    metrics(y_te, p, threshold = th) |> select(precision, recall, f1)
  )
}

common_topics <- topic_counts |> filter(n >= 80) |> pull(topics)
common_topics <- setdiff(common_topics, "Other")

results <- map_dfr(common_topics, fit_topic) |> arrange(desc(auc))
results |> mutate(across(where(is.numeric), \(x) round(x, 2)))
```

```{r}
#| label: fig-auc
#| fig-cap: "Test-set AUC for each topic. The dashed line is chance."

ggplot(results, aes(auc, reorder(topic, auc))) +
  geom_point(size = 3, color = vcu_teal) +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  xlim(0.45, 1) +
  labs(x = "AUC on held-out students", y = NULL)
```

In this synthetic data, how well a topic is predicted mostly reflects how many examples it has and how much noise was built in, not how abstract the topic is. With real interviews you would expect a wider spread: concrete topics such as housing or financial aid tend to be easy, because students use distinctive words for them, while abstract topics such as "sense of belonging" tend to be harder, because students express them in many different ways. When that happens, a topic's difficulty tells you something about the topic as much as about the model.

Two cautions on reading this table. First, each topic is scored on a single test split with only 20 to 75 true cases, so differences of a few hundredths in AUC are noise. Second, you may see a `glmnet` convergence warning on some topics. It means the optimizer did not finish for the very smallest penalties on a weak-signal topic, and `glmnet` returns the solutions for the larger penalties instead. It is worth knowing about but does not invalidate the table.

## What supervised learning gives you and what it costs

| | Unsupervised (Ch. 3-6) | Supervised (this chapter) |
|---|---|---|
| Needs labels | No | Yes |
| Answers the question | "What patterns are in here?" | "Can we reproduce these codes?" |
| Measures success against | Your judgment | A held-out answer key |
| Biggest risk | Seeing patterns that are not there | Learning the labelers' habits and quirks |

A good model does not prove the labels are *right*. It shows that the labels are *learnable*: that the text contains enough signal for a computer to reproduce what the interviewers did. That is a useful result, but it is a smaller claim than "the model understands student concerns."

## Summary

- Split first, by student, and build the vocabulary from training data only.
- A lasso on TF-IDF gives a strong, interpretable baseline.
- Judge with precision, recall, and AUC, not accuracy, when classes are imbalanced.
- Read the errors. They tell you whether to improve the model or the labels.
